Modelling the emergence of whisker barrels

### 2020-07-06 paper/briefpaper/brief.tex
Finished editting (I think) but note final fixme in movie caption still

Spatial patterns in neural connectivity provide clues about the constraints
under which brains evolve and develop \citep{purves_iterated_1992}. Perhaps
the most distinctive pattern can be found in the barrel cortex of many rodent
species \citep{woolsey_structural_1970}. The barrels are identifiable soon
after birth in layer 4 of primary somatosensory cortex as dense clusters of
thalamocortical axons, which are enclosed by borders a few neurons thick from
postnatal day 3 \citep{erzurumlu_development_2012}.
%
% These timings, from erzurumlu_development_2012 originate from
% sehara_neuronal_2011, where they relate to mouse.
\mpsix{In the mouse, axons from the trigeminal ganglion arrive in the principal sensory
  nucleus of the trigeminal nerve (PrV) at E12, then axons from the PrV arrive
  in the ventroposteromedial nucleus of the thalamus (VPM) at E17, then axons
  from the VPM arrive in the cortical plate at E18/P0. Distinct whisker-related
  clusters then become apparent in the PrV at P0-P1, in the VPM at P2-P3, and
  in the cortex at P3-P5} \citep{erzurumlu_development_2012,sehara_neuronal_2011}.
%
In the plane tangential to
the cortical surface the barrels constitute a somatotopic map of the whiskers,
with cells within adjacent barrels responding most strongly and quickly to
deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. Barrel
patterning reflects subcortical whisker maps comprising cell aggregates called
barrelettes in the brainstem and barreloids in the thalamus
\citep{ma_barrelettesarchitectonic_1991,van_der_loos_barreloids_1976}.

Barrel formation requires afferent input from whisker stimulation and thalamic
calcium waves \citep{anton-bolanos_prenatal_2019}, and depends on a complex
network of axon guidance molecules such as ephrin-A5 and A7 and adhesion
molecules such as cadherin-6 and 8
\citep{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}.  This network
is orchestrated by interactions between morphogens Fgf8 and Fgf17 and
transcription factors Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed
in gradients \mpfour{across the full cortical sheet} that mark orthogonal axes
and can be manipulated to stretch,
shrink, shift, and even duplicate barrels
\citep{assimacopoulos_fibroblast_2012}.

The barrel boundaries form a Voronoi tessellation across the cortical sheet
\citep{senft_mouse_1991} (Fig.\,\ref{fig:main}A), suggesting that barreloid topology is
preserved in the projection of thalamocortical axons into the cortex, and that
a barrel forms by lateral axon branching from an initial center-point that
ceases upon contact with axons branching from adjacent centers.  However, the
assumption of pre-arranged center-points is difficult to resolve with the
observation that axons arrive in the cortical plate as an undifferentiated
bundle, \emph{prior} to barreloid formation \citep{agmon_organized_1993}.

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the density of
connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a
1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by
$i$. The model was derived from the assumption that the rates at which $a_i$
and $c_i$ grow are reciprocally coupled. Extending the original 1D model to
simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and
$c_i(\mb{x},t)$, and model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \MPtwo{+ \chi_i}\right) - \frac{\partial c_i}{\partial t}.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. Note that computing the divergence in
simulation requires cells on the cortical sheet to communicate with immedately
adjacent cells only (see \emph{Materials \& Methods}). The second term on the
right quantifies the coupling between axon branching and synaptogenesis. Here
  $\chi_i=0$ is a placeholder.

@@

\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig1.png}
    \caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
      by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
      measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
      perfect Voronoi tesselation. Right shows the initial distribution of axon
      branching density ($a$) for one thalamocortical projection, and two
      molecular guidance fields ($\rho$), \cmnt{where the domain $S$ has been
        traced from \textbf{A}}. \textbf{B} The strengths of interaction
      $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
      projections by the lengths of green and blue arrows respectively, assuming
      that similar fields aligned to the posterior-anterior and medial-lateral
      axes in the ventroposterior medial nucleus of the thalamus are sampled at
      the locations of putative barreloid centers (reconstructed from
      \citealp{haidarliu_size_2001}, \MPone{Fig.\,5b}). \textbf{C} Simulation results for parameters
      \MPtwo{$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\gamma\in\pm 2$,
      $\epsilon=1.2$} and $\delta{t}=0.0001$. Colours indicate the thalamic
      projection for which the connection density is maximal and black lines
      delineate boundaries (see Movie
      S1). \textbf{D} Red dots show the Honda metric, \metrics{$\delta(t)$}, obtained from \metrics{the}
      simulation approaching that obtained from \metrics{real} barrels in \textbf{A} (dotted
      line); black squares \metrics{provide a} measure \metrics{of} the correspondence
      between the real and
      simulated barrel shapes, $\eta(t)$ (the \metrics{pattern quality metric---see
        Materials \& Methods; units mm$^3$});
      \metrics{grey hexagons \metrics{provide a measure of how selectively each
          cortical site is innervated};
        $\omega(t) = \oiint_{S} \mu(\mb{x},t) \mathrm{d}S$, where
        $\mu(\mb{x}) \equiv \mathrm{max}_i\big(c_i(\mb{x},t)\big)\big/\sum_{j=1}^{N} c_j(\mb{x},t)$.}
      \textbf{E} \metrics{Plotted across the cortical sheet, the selectivity
        develops to reveal an alignment with the emergent barrel boundary
        shapes}. \metrics{The colour map
        indicates the values of $\mu(\mb{x})$. Overlaid
        contours show $c_i > 0.95 \mathrm{max}_i\big(c_i(\mb{x})\big)$.} All scale bars 1\,mm.}
    \label{fig:main}
  \end{fullwidth}
\end{figure}

@@

\mpfive{We reasoned that an arbitrary number of distinct field locations may
  be specified using a minimum of two guidance fields, if the tendency for the
  density of each projection to concentrate is appropriately balanced against
  the tendency of projections that interact more strongly with the guidance
  fields to migrate further in the directions specified by their
  gradients. Accordingly, projections that interact most strongly with a given
  guidance gradient would come to occupy cortical locations at which that
  field has extreme values, leaving adjacent locations available to be
  occupied by projections with the next strongest interactions, and so
  forth. This would in principle allow the \emph{relative} locations of the
  fields to be specified by the relative values of the interaction parameters,
  $\gamma$, and hence for a topological map in the cortex to be specified by a
  spatial ordering of the $\gamma$ values at the level of the thalamus.}

@@

\MPone{In addition, the model we have outlined requires that molecular
  guidance gradients in the cortex are complemented by graded values of the
  interaction strengths, $\gamma$, at the level of the thalamus. While the
  precise mechanisms by which thalamic and cortical gradients interact during
  development have not been fully characterised, the presence of complementary
  thalamic and cortical molecular guidance gradients has been well established
  experimentally. In particular, the EphA4 receptor and its ligand ephrin-A5
  are distributed in complementary gradients in the somatosensory thalamus and
  cortex} (\citealp{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}).
\MPone{Cells originating in the medial part of ventrobasal thalamus
  (comprising VPM and VPL) express high levels of EphA receptors and project
  to the lateral part of S1, which expresses low levels of ephrin-A5, and
  cells originating in the lateral part of ventrobasal thalamus express low
  levels of EphA receptors and project to the medial part of S1, which
  expresses high levels of ephrin-A5} (see
\citealp{gao_regulation_1998,dufour_area_2003,vanderhaeghen_developmental_2004,speer_grading_2005,torii_role_2013}).
\MPone{We assume that such patterning arises because relative strengths of
  interaction with guidance molecules (e.g., ephrin-A5) in the cortex are
  correlated with the relative concentrations of complementary molecules
  (e.g., EphA4) in the thalamus, and thus with thalamic position along the
  axis to which their gradients are aligned.}

@@

A cortical boundary enclosing \cmnt{the} 41 \cmnt{macrovibrissae} barrels was
traced from a cytochrome oxidase stain from \cite{zheng_signal_2001}
\cmnt{(using original data kindly supplied by the authors)}, and
Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved for $N=41$ projections on the
resulting domain, $S$, using $M=2$ linear signalling gradients aligned with
the anterior-posterior and medial-lateral axes.  \mpfour{These gradients are
  shown with the barrel field boundary in Fig.\,\ref{fig:main}A for clarity,
  though like ephrin-5 they may be thought of as extending across the cortical
  hemisphere \citep{miller_epha7-ephrin-a5_2006}. Simulations were stepped
  through 30000 iterations of Eqs.\,\ref{eq:dc}--\ref{eq:comp} ($\delta
  t=0.0001$).}

\cmnt{Across a wide range of parameter values, random initial conditions (a
  uniform random distribution for $a(\mb{x},0)\in(0.2,0.4)$, $c(\mb{x},0)=0$)
  eventually yielded a clear Voronoi-like tessellation of topographically
  organized thalamocortical projections, confirming that barrel maps can
  self-organize in the absence of pre-specified center points. The
  organization is apparent in a plot of the identity of the projection for
  which the connection density is maximal at each simulated cortical location,
  as shown in Fig.\,\ref{fig:main}C. Parameters for the example simulation
  shown in Fig.\,\ref{fig:main}C (see also Movie S1) were obtained by
  conducting a full parameter sweep and choosing a combination of parameter
  values ($\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\epsilon=1.2$) that
  scored well against the following three measures.}

@@

% Full page width
\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig2.png}
    \caption{\MPthreePar{\textbf{A} Exploring the parameter space of the
        model given in Eqs.\,\ref{eq:dc}--\ref{eq:comp}. Graphs show the three
        pattern metrics for $t=30000$ steps as colourmaps: Honda $\delta$ (top
        row) shows conformance to a Voronoi pattern; $\eta$ (middle row) gives
        the simulated/experimental barrel area difference and $\omega$ (bottom
        row) is a measure of selectivity of the connection densities for the
        41 thalamocortical axon types. The colour maps are chosen so that for
        each row, lighter (orange and yellow) colours indicate desirable
        properties for the simulations: Lower values of $\delta$ are yellow,
        indicating good Voroni patterns; Lower values of $\eta$ indicate good
        pattern matching; Higher values of $\omega$ indicate greater
        selectivity. White squares indicate that the simulation was not stable
        for that parameter set.
%
        The parameter space explored is three dimensional with the parameter
        ratio $\alpha/\beta$ varied along the columns of graphs, the
        competition parameter $\epsilon$ varied on the $y$-axes and the
        diffusion constant, $D$ varied along the $x$-axes. An asterisk (*)
        marks the parameter set displayed in Fig.\,\ref{fig:main}. Boxes i),
        ii)~and iii)~mark parameter sets for which corresponding patterns are
        shown in \textbf{B} (for $t=30000$ steps).
%
        \textbf{B} The effect of varying the diffusion constant,
        $D$. i)~Increasing $D$ above the optimum value gives an expansion of
        the outer barrels and a corresponding compression of the barrels
        within. ii)~Conversely, an expansion of the central barrels and
        compression of those near the boundary results from reducing $D$ from
        its optimum. iii) Reducing the diffusion constant is equivalent to
        increasing the size of the domain $S$ and so this simulation, in which
        $D$ had the lowest stable value, can be thought to represent an animal
        with a large cortical area but with unchanged cortical complexity. A
        large proportion of the domain is occupied by the projections with
        intermediate interaction parameters; those with strong interaction
        parameters are compressed around the edge of the domain. A barrel
        pattern fails to form.}}
    \label{fig:paramsweep}
  \end{fullwidth}
\end{figure}

@@

% Full page width
\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig3.png}
    \caption{
      \MPthreeSens{\textbf{A} Sensitivity of the system pattern
        metrics $\delta(t)$ (top row), $\omega(t)$ (middle row) and $\eta(t)$
        (bottom row), evaluated at $t=30000$ steps,
        to three classes of perturbation. $y$-axes and colour scales have identical ranges
        to the colour scales in Fig.\,\ref{fig:paramsweep}, to ease comparison
        with the parameter exploration.
        % $a(\mb{x},0)\in(0.2,0.4)$
        \emph{Left column:} Random noise drawn from a uniform distribution,
        $(\gamma_{max}-\gamma_{min})\,U(0,\nu_\gamma)$, added
        to the values for the interaction parameters, $\gamma_{i,j}$ used in
        Fig.\,\ref{fig:main}.
        %
        \emph{Middle column:} Random noise $(\rho_{max} - \rho_{min})\,U(0, \nu_{\rho})$
        was added to each hex element of $\rho_1(\mb{x})$ and $\rho_2(\mb{x})$ and the result
        was smoothed by convolution with a symmetric 2D
        Gaussian kernel of width $\sigma_\rho$. Colour maps in this column
        thus show the effect of changing the magnitude and the length scale of
        noise in the guidance fields.
        %
        \emph{Right column:} Colour maps show the result of setting the rotational angle
        of the linearly varying guidance field, $\rho_1(\mb{x})$, to $\phi_{\rho_1}$,
        and modifying its overall gain to $G_{\rho_1}$, whilst keeping the
        parameters of $\rho_2(\mb{x})$
        unchanged from those used in the example simulation, for which
        $\phi_{\rho_2}=84$\textdegree and $G_{\rho_2}=1$.
        %
        \textbf{B} Four ways in which the perturbations in \textbf{A} affect
        the patterns. i) The effect of significant interaction parameter noise
        causes barrels to become swapped from the positions they would
        otherwise take. ii) High magnitude, short lengthscale noise in
        $\rho(\mb{x})$ leads to meandering boundaries between barrels. iii)
        Shows the effect of reducing the slope of $\rho_1$ (by a factor of 10)
        which reduces the strength of the axons' interaction with the guidance
        field in the left-right sense. Barrel rows B, C and D become `crushed'
        down the center line, and edge barrels dominate. iv) The effect of
        rotating $\rho_1$ by 20\textdegree~is a relatively mild distortion of
        the pattern in the anticlockwise sense.
      }
    }
    \label{fig:sens}
  \end{fullwidth}
\end{figure}

@@

\MPthreePred{To further investigate the interplay of genetic and intrinsic
  mechanisms we simulated two well known experimental manipulations of the
  whisker barrel system. First, we simulated a seminal barrel duplication
  experimental paradigm}
\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012}
\MPthreePred{in which the growth factor Fgf8, which is normally expressed at
  the anterior end of the cortical subplate from around E9.5}
\citep{crossley_mouse_1995}, \MPthreePred{is ectopically expressed (by
  electroporation) also at the posterior pole of a cortical
  hemisphere. Because Fgf8 sets off a chain of gene interactions involved in
  controlling the graded expression of ephrin and cadherin molecules, its
  misexpression at the posterior pole could be expected to mirror those
  gradients which usually vary along the rostrocaudal axis. We assume that
  this results in a mirror of the primary barrel cortex boundary along the
  rostrocaudal axis, to match the experimental results of}
\cite{assimacopoulos_fibroblast_2012}. \MPthreePred{We also assume that a
  further consequence of the posterior Fgf8 electroporation is the mirrored
  anterior-posterior guidance gradient $\rho_1$ (Fig.\,\ref{fig:fgf8}A). With
  the modified $\rho_1$ guidance gradient the simulation was run with $N=41$
  thalamocortical axon types, using the parameters of the example
  simulation. The result, after 30000 iterations, was two mirror-symmetrical
  barrel fields comprising $2N$ barrels (Fig.\,\ref{fig:fgf8}B), consistent
  with the outcome of the original experiments.}

>>>>

Spatial patterns in neural connectivity provide clues about the constraints
under which brains evolve and develop \citep{purves_iterated_1992}. Perhaps
the most distinctive pattern can be found in the barrel cortex of many rodent
species \citep{woolsey_structural_1970}. The barrels are identifiable soon
after birth in layer 4 of primary somatosensory cortex as dense clusters of
thalamocortical axons, which are enclosed by borders a few neurons thick from
postnatal day 3 \citep{erzurumlu_development_2012}.
%
In the plane tangential to
the cortical surface the barrels constitute a somatotopic map of the whiskers,
with cells within adjacent barrels responding most strongly and quickly to
deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. Barrel
patterning reflects subcortical whisker maps comprising cell aggregates called
barrelettes in the brainstem and barreloids in the thalamus
\citep{ma_barrelettesarchitectonic_1991,van_der_loos_barreloids_1976}.

Barrel formation requires afferent input from whisker stimulation and thalamic
calcium waves \citep{anton-bolanos_prenatal_2019}, and depends on a complex
network of axon guidance molecules such as ephrin-A5 and A7 and adhesion
molecules such as cadherin-6 and 8
\citep{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}.  This network
is orchestrated by interactions between morphogens Fgf8 and Fgf17 and
transcription factors Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed
in gradients \mpfour{spanning the cortical sheet} that mark orthogonal axes
and can be manipulated to stretch,
shrink, shift, and even duplicate barrels
\citep{assimacopoulos_fibroblast_2012}.

The barrel boundaries form a Voronoi tessellation
\citep{senft_mouse_1991} (Fig.\,\ref{fig:main}A), suggesting that barreloid topology is
preserved in the projection of thalamocortical axons into the cortex, and that
a barrel forms by lateral axon branching from an initial center-point that
ceases upon contact with axons branching from adjacent centers.  However, the
assumption of pre-arranged center-points is difficult to resolve with the
observation that axons arrive in the cortical plate as an undifferentiated
bundle, \emph{prior} to barreloid formation \citep{agmon_organized_1993}.
\mpsix{In mice, axons from the trigeminal ganglion arrive in the principal sensory
  nucleus of the trigeminal nerve (PrV) at E12, then axons from the PrV arrive
  in the ventroposteromedial nucleus of the thalamus (VPM) at E17, then axons
  from the VPM arrive in the cortical plate at E18/P0. Distinct whisker-related
  clusters then become apparent in the PrV at P0-P1, in the VPM at P2-P3, and
  in the cortex at P3-P5} \citep{erzurumlu_development_2012,sehara_neuronal_2011}.

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the density of
connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a
1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by
$i$. The model was derived from the assumption that the rates at which $a_i$
and $c_i$ grow are reciprocally coupled. Extending the original 1D model to
simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and
$c_i(\mb{x},t)$, and model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \MPtwo{+ \chi_i}\right) - \frac{\partial c_i}{\partial t}.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. Note that computing the divergence in
simulation requires cells on the cortical sheet to communicate with immedately
adjacent cells only (see \emph{Materials \& Methods}). The second term on the
right represents the coupling between axon branching and synaptogenesis. Here
  $\chi_i=0$ is a placeholder.

@@

\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig1.png}
    \caption{The emergence of whisker barrels. \textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
      by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
      measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
      perfect Voronoi tesselation. Right shows the initial distribution of axon
      branching density ($a$) for one thalamocortical projection, and two
      molecular guidance fields ($\rho$), \cmnt{where the domain $S$ has been
        traced from \textbf{A}}. \textbf{B} The strengths of interaction
      $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
      projections by the lengths of green and blue arrows respectively, assuming
      that similar fields aligned to the posterior-anterior and medial-lateral
      axes in the ventroposterior medial nucleus of the thalamus are sampled at
      the locations of putative barreloid centers (reconstructed from
      \citealp{haidarliu_size_2001}, their \MPone{Fig.\,5b}). \textbf{C} Results for the example simulation, with parameters
      \MPtwo{$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\gamma\in\pm 2$,
      $\epsilon=1.2$} and $\delta{t}=0.0001$. Colours indicate the thalamic
      projection for which the connection density is maximal and black lines
      delineate boundaries (see Movie
      S1). \textbf{D} Red dots show the \emph{Honda}--\metrics{$\delta(t)$} metric obtained from \metrics{the}
      simulation approaching that obtained from the \metrics{real} barrels in \textbf{A} (dotted
      line); black squares show the \emph{pattern difference} metric $\eta(t)$, and reveal the emergence of a correspondence between the real and simulated barrel shapes (units mm$^3$); \metrics{grey hexagons show \metrics{how selectively each cortical site is innervated};
        $\omega(t) = \oiint_{S} \mu(\mb{x},t) \mathrm{d}S$, where
        $\mu(\mb{x}) \equiv \mathrm{max}_i\big(c_i(\mb{x},t)\big)\big/\sum_{j=1}^{N} c_j(\mb{x},t)$.}
      \textbf{E} \metrics{Plotted across the cortical sheet, the selectivity
        develops to reveal an alignment with the emergent barrel boundary
        shapes}. \metrics{Greyscale colour indicates values of $\mu(\mb{x})$ and overlaid
        contours show $c_i > 0.95 \mathrm{max}_i\big(c_i(\mb{x})\big)$.} All scale bars 1\,mm.}
    \label{fig:main}
  \end{fullwidth}
\end{figure}

@@

\mpfive{We reasoned that an arbitrary number of distinct field locations may
  be determined by a minimum of two guidance fields, if a tendency for the
  density of projections to concentrate is appropriately balanced with
  the tendency of projections that interact more strongly with the guidance
  fields to migrate further in the directions specified by their
  gradients. Accordingly, projections that interact most strongly with a given
  guidance gradient would come to occupy cortical locations at which that
  field has extreme values, leaving adjacent locations available to be
  occupied by projections with the next strongest interactions, and so
  forth. This would in principle allow the \emph{relative} locations of the
  fields to be specified by the relative values of the interaction parameters,
  $\gamma$, and hence for a topological map in the cortex to be specified by a
  spatial ordering of the $\gamma$ values at the level of the thalamus.}

@@

\MPone{In addition, the model we have outlined requires that molecular
  guidance gradients in the cortex are complemented by graded values of the
  interaction strengths, $\gamma$, at the level of the thalamus. While the
  precise mechanisms by which thalamic and cortical gradients interact during
  development have not been fully characterised, the presence of complementary
  thalamic and cortical guidance gradients has been well established
  experimentally. In particular, the EphA4 receptor and its ligand ephrin-A5
  are distributed in complementary gradients in the somatosensory thalamus and
  cortex} (\citealp{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}).
\MPone{Cells originating in the medial part of ventrobasal thalamus
  (comprising VPM and VPL) express high levels of EphA receptors and project
  to the lateral part of S1, which expresses low levels of ephrin-A5, and
  cells originating in the lateral part of ventrobasal thalamus express low
  levels of EphA receptors and project to the medial part of S1, which
  expresses high levels of ephrin-A5} (see
\citealp{gao_regulation_1998,dufour_area_2003,vanderhaeghen_developmental_2004,speer_grading_2005,torii_role_2013}).
\MPone{We assume that such patterning arises because the relative strengths of
  interaction with guidance molecules (e.g., ephrin-A5) in the cortex are
  correlated with the relative concentrations of complementary molecules
  (e.g., EphA4) in the thalamus, and thus with thalamic position along the
  axis to which their gradients are aligned.}

@@

A cortical boundary enclosing barrels for 41 \cmnt{macrovibrissae} was
traced from a cytochrome oxidase stain from \cite{zheng_signal_2001}
\cmnt{(using original data kindly supplied by the authors)}, and
Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved for $N=41$ projections on the
resulting domain, $S$, using $M=2$ linear signalling gradients aligned with
the anterior-posterior and medial-lateral axes.  \mpfour{These gradients are
  shown with the barrel field boundary in Fig.\,\ref{fig:main}A for clarity,
  though like ephrin-5 they may be thought of as extending across the cortical
  hemisphere \citep{miller_epha7-ephrin-a5_2006}. Simulations were stepped
  through 30000 iterations of Eqs.\,\ref{eq:dc}--\ref{eq:comp} ($\delta
  t=0.0001$).}

\cmnt{Across a wide range of parameter values, random initial conditions (a
  uniform random distribution for $a(\mb{x},0)\in(0.2,0.4)$, $c(\mb{x},0)=0$)
  eventually yielded a clear Voronoi-like tessellation of topographically
  organized thalamocortical projections, confirming that barrel maps can
  self-organize in the absence of pre-specified center points. The
  organization is apparent in a plot of the identity of the projection for
  which the connection density is maximal at each simulated cortical location,
  as shown in Fig.\,\ref{fig:main}C. Parameter values for this example simulation
  (see also Movie S1) were obtained by
  conducting a full parameter sweep and choosing a combination ($\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\epsilon=1.2$) that
  scored well against the following three measures.}

@@

% Full page width
\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig2.png}
    \caption{\MPthreePar{Exploring the parameter space. \textbf{A} Colour indicates the quality of the pattern at $t=30000$ steps, against three measures: Low values of \emph{Honda}--$\delta$ (top
        row) suggest a Voronoi-like pattern of fields. The \emph{pattern difference}, $\eta$ (middle row) measures the difference in the area and arrangement of barrels in real and simulated fields. The \emph{selectivity}, $\omega$ (bottom
        row), measures the specificity with which the cortical sheet is innervated. Colour maps are chosen so that lighter (orange and yellow) colours indicate higher quality patterns against each measure. White squares indicate combinations of parameters for which simulations were numerically unstable.
%
        The parameter space explored is three dimensional with the ratio $\alpha/\beta$ varying between plots, and the
        competition parameter $\epsilon$ and the diffusion constant $D$ varying within plots. An asterisk (*)
        marks the parameters used in Fig.\,\ref{fig:main}. Boxes i),
        ii)~and iii)~mark parameter sets for which corresponding patterns are
        shown in \textbf{B} (for $t=30000$ steps).
%
        \textbf{B} Varying the diffusion constant $D$ generates qualitatively different patterns. Higher values 
        cause expansions of the peripheral barrels and a corresponding compression of the inner barrels (i). Lower values instead cause an expansion of the central barrels and compression of the peripheral barrels (ii). Further reducing the rate of diffusion (iii), which is equivalent to increasing the size of the domain and hence simulating development in an animal with a larger cortex, causes a large area to be occupied by projections with intermediate interaction parameters; those with strong interaction parameters are compressed around the edge of the domain, and consequently a barrel pattern fails to form.}}
    \label{fig:paramsweep}
  \end{fullwidth}
\end{figure}

@@

% Full page width
\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig3.png}
    \caption{
      \MPthreeSens{Sensitivity analysis. \textbf{A} Metrics of map quality $\delta(t)$ (top row), $\omega(t)$ (middle row) and $\eta(t)$ (bottom row), were evaluated at $t=30000$ steps. The $y$-axes and colour scales have identical ranges
        to the colour scales in Fig.\,\ref{fig:paramsweep}, for easy comparison. \emph{Left column:} The effect of adding noise drawn from a uniform distribution,
        $(\gamma_{max}-\gamma_{min})\,U(0,\nu_\gamma)$,
        to the values of the interaction parameters, $\gamma$ used in
        Fig.\,\ref{fig:main}.
        %
        \emph{Middle column:} The effect of changing the magnitude and the length scale of
        noise applied to the guidance fields. Uniform random noise $(\rho_{max} - \rho_{min})\,U(0, \nu_{\rho})$ was added
        to each (hexagonal) element of $\rho_1(\mb{x})$ and $\rho_2(\mb{x})$ and the result
        was smoothed by convolution with a symmetric 2D
        Gaussian kernel of width $\sigma_\rho$. 
        %
        \emph{Right column:} The effect of setting the rotational angle
        of the linearly varying guidance field, $\rho_1(\mb{x})$, to $\phi_{\rho_1}$,
        and modifying its overall gain to $G_{\rho_1}$, whilst keeping the
        parameters of $\rho_2(\mb{x})$
        unchanged from those used in the example simulation, for which
        $\phi_{\rho_2}=84$\textdegree and $G_{\rho_2}=1$.
        %
        \textbf{B} Four ways in which the perturbations in \textbf{A} affect
        the patterns. i) The effect of significant interaction parameter noise
        results in topological defects. ii) High magnitude, short lengthscale noise in
        $\rho(\mb{x})$ leads to non-straight edges between adjacent barrels. iii)
 Reducing the slope of gradient $\rho_1$ (by a factor of 10)
causes barrel rows B, C and D to become `crushed'
        down the center line, and edge barrels to dominate. iv) Rotating $\rho_1$ by 20\textdegree~causes a slight distortion of the pattern, resulting in an overall anticlockwise rotation of the field locations.
      }
    }
    \label{fig:sens}
  \end{fullwidth}
\end{figure}

@@

\MPthreePred{To further investigate the interplay of genetic and intrinsic
  mechanisms we simulated two well known experimental manipulations of 
barrel development. First, we simulated a seminal barrel duplication
  paradigm}
\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012}
\MPthreePred{in which the growth factor Fgf8, which is normally expressed at
  the anterior end of the cortical subplate from around E9.5}
\citep{crossley_mouse_1995}, \MPthreePred{is ectopically expressed (by
  electroporation) also at the posterior pole. We assume that
  this results in a mirror of the primary barrel cortex boundary along the
  rostrocaudal axis}
\cite{assimacopoulos_fibroblast_2012} \MPthreePred{and a the mirroring of the
  anterior-posterior guidance gradient $\rho_1$ at the border between them (Fig.\,\ref{fig:fgf8}A). 
  The result after 30000 iterations, and otherwise using the parameters of the example simulation, was two mirror-symmetrical
  barrel fields comprising $2N$ barrels (Fig.\,\ref{fig:fgf8}B), consistent
  with the outcome of the original experiments.}

% This figure is text width
\begin{center}
  \includegraphics[width=\linewidth]{./Fig4.png} \captionof{figure}{
    \MPthreePred{Simulating altered barrel development.  Guidance fields
      (\textbf{A}) and emergent barrel pattern (\textbf{B}) in a Fgf8
      misexpression experiment
      (c.f.~\citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
      $\rho_1$ from Fig.\,\ref{fig:main}A at the join of the original boundary
      with its mirror. All other model parameters match those in
      Fig.\,\ref{fig:main}C. \textbf{C} Simulating
      whisker trimming by reducing the competitiveness,
      $\epsilon$, of the one projection. For C3 only, $\epsilon$ was mutiplied by
      $m\in(0,1)$. The pattern shown is that formed
      after 30000 steps with $m=0.86$, which reduces the size of the C3 field to 65\% of its original size,
      matching the average barrel area reduction observed by }
    \citealp{kossut_effects_1992}. \MPthreePred{\textbf{D} The area of the C3 barrel
reduces as $m$ is reduced, whereas
      the mean area of neighbouring barrels (B3, C2, C4, D2 \& D3, grey
      circles) increases. The dotted grey line indicates $m=0.86$ for
      comparison with panel C.}}
  \label{fig:fgf8}
\end{center}
